### 2025-06-30 paper/quarto_aaai/paper.tex
Fix some awkward references after reformatting for AAAI

The first challenge has attracted an abundance of work on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations, aiming to mitigate scenarios in which
practitioners deploy opaque models and have to blindly rely on their
predictions. On many occasions, this has happened in practice, causing
harms to people who were adversely and unfairly affected by automated
decision-making (ADM) systems involving opaque models (see, e.g.,
\citet{oneil2016weapons}). Effective XAI tools can also aid in
monitoring models and providing recourse, empowering people to turn
negative outcomes (e.g., ``loan application rejected'') into positive
ones (e.g., ``loan application accepted''). In line with this, our work
builds upon \textbf{counterfactual explanations} (CE) proposed by
\citet{wachter2017counterfactual}; CEs prescribe minimal changes for
factual inputs that, if implemented, would prompt some fitted model to
produce an alternative, more desirable output.

@@

\citet{grathwohl2020your} observed that any classifier can be
re-interpreted as a joint energy-based model that learns to discriminate
output classes conditional on the observed (training) samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). \citet{grathwohl2020your} show that JEMs can
be trained to perform well at both tasks by directly maximizing the
joint log-likelihood:
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\),
where the first term can be optimized using cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\), they
minimize the contrastive divergence between the observed samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\).

@@

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \citet{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we choose to work with a different distinction between CE and AE
than \citet{freiesleben2022intriguing} who consider misclassification as
the key distinguishing feature of AE. One of the key observations of
this work is that we can leverage CEs during training and get
adversarial examples essentially for free, which can be used to reap the
aforementioned benefits of adversarial training.

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). These datasets are generated using the library of
\citep{altmeyer2023explaining} and we present them in the supplementary
appendix. Next, we have four real-world binary tabular datasets from the
domain of economics: \emph{Adult} (a.k.a. Census data) of
\citep{becker1996adult2}, California housing (\emph{CH}) of
\citep{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
\citep{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) of
\citep{kaggle2011give2}. Finally, for the convenience of illustration,
we use of the 10-class \emph{MNIST} vision dataset
\citep{lecun1998mnist}.

@@

\emph{Interventions on features have implications for fairness.} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features. Model owners can use our
solution to support the fair and equitable treatment of decision
subjects, but they could also misuse it by enforcing explanations based
on features that are more difficult to modify by some (group of)
individuals. When used irresponsibly, CT could result in an unfairly
assigned burden of recourse \citep{sharma2020certifai}, threatening the
equality of opportunity \citep{bell2024fairness}. Additionally, CT
requires mutability constraints for the features considered by the
model. Even if all immutable features are protected, there may exist
proxies that are mutable, and hence should not be protected, but
preserve sufficient information about the principals to hinder the
protections. Deciding on actionability is still a major open challenge
in the AR literature \citep{venkatasubramanian2020philosophical}
impacting the capacity of CT to fulfill its intended goal.

>>>>

The first challenge has attracted an abundance of work on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations, aiming to mitigate scenarios in which
practitioners deploy opaque models and have to blindly rely on their
predictions. On many occasions, this has happened in practice, causing
harms to people who were adversely and unfairly affected by automated
decision-making (ADM) systems involving opaque models; see, e.g.,
\citet{oneil2016weapons}. Effective XAI tools can also aid in monitoring
models and providing recourse, empowering people to turn negative
outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
``loan application accepted''). In line with this, our work builds upon
\textbf{counterfactual explanations} (CE) proposed by
\citet{wachter2017counterfactual}; CEs prescribe minimal changes for
factual inputs that, if implemented, would prompt some fitted model to
produce an alternative, more desirable output.

@@

\citet{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model that learns to discriminate
output classes conditional on the observed (training) samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood:
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\),
where the first term can be optimized using cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\), they
minimize the contrastive divergence between the observed samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\).

@@

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \citet{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we work with a different distinction between CE and AE than
\citet{freiesleben2022intriguing} who considers misclassification as the
distinguishing feature of adversarial examples. One of the key
observations of this work is that we can leverage CEs during training
and get AEs essentially for free to reap the aforementioned benefits of
adversarial training.

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). These datasets are generated using the library of
\citet{altmeyer2023explaining} and we present them in the supplementary
appendix. Next, we have four real-world binary tabular datasets from the
domain of economics: \emph{Adult} (a.k.a. Census data) of
\citet{becker1996adult2}, California housing (\emph{CH}) of
\citet{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
\citet{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) from
\citet{kaggle2011give2}. Finally, for the convenience of illustration,
we use the 10-class \emph{MNIST} dataset \citep{lecun1998mnist}.

@@

\emph{Interventions on features have implications for fairness.} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features. Model owners can use our
solution to support the fair and equitable treatment of decision
subjects, but they could also misuse it by enforcing explanations based
on features that are more difficult to modify by some (group of)
individuals. When used irresponsibly, CT could result in an unfairly
assigned burden of recourse \citep{sharma2020certifai}, threatening the
equality of opportunity \citep{bell2024fairness}. Additionally, CT
requires mutability constraints for the features considered by the
model. Even if all immutable features are protected, there may exist
proxies that are mutable, and hence should not be protected, but
preserve sufficient information about the principals to hinder these
protections. Deciding on actionability is still a major open challenge
in the AR literature \citep{venkatasubramanian2020philosophical}
impacting the capacity of CT to fulfill its intended goal.

### 2025-06-30 paper/sections/discussion.qmd
Fix some awkward references after reformatting for AAAI

*Interventions on features have implications for fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features. Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. When used irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity [@bell2024fairness]. Additionally, CT requires mutability constraints for the features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable, and hence should not be protected, but preserve sufficient information about the principals to hinder the protections. Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.

>>>>

*Interventions on features have implications for fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features. Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. When used irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity [@bell2024fairness]. Additionally, CT requires mutability constraints for the features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable, and hence should not be protected, but preserve sufficient information about the principals to hinder these protections. Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.

### 2025-06-30 paper/sections/experiments.qmd
Fix some awkward references after reformatting for AAAI

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). These datasets are generated using the library of [@altmeyer2023explaining] and we present them in the supplementary appendix. Next, we have four real-world binary tabular datasets from the domain of economics: *Adult* (a.k.a. Census data) of [@becker1996adult2], California housing (*CH*) of [@pace1997sparse], Default of Credit Card Clients (*Cred*) of [@yeh2016default], and Give Me Some Credit (*GMSC*) of [@kaggle2011give2]. Finally, for the convenience of illustration, we use of the 10-class *MNIST* vision dataset [@lecun1998mnist].

>>>>

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). These datasets are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix. Next, we have four real-world binary tabular datasets from the domain of economics: *Adult* (a.k.a. Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2. Finally, for the convenience of illustration, we use the 10-class *MNIST* dataset [@lecun1998mnist].

### 2025-06-30 paper/sections/intro.qmd
Fix some awkward references after reformatting for AAAI

The first challenge has attracted an abundance of work on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations, aiming to mitigate scenarios in which practitioners deploy opaque models and have to blindly rely on their predictions. On many occasions, this has happened in practice, causing harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models (see, e.g., @oneil2016weapons). Effective XAI tools can also aid in monitoring models and providing recourse, empowering people to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "loan application accepted"). In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual; CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.

>>>>

The first challenge has attracted an abundance of work on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations, aiming to mitigate scenarios in which practitioners deploy opaque models and have to blindly rely on their predictions. On many occasions, this has happened in practice, causing harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models; see, e.g., @oneil2016weapons. Effective XAI tools can also aid in monitoring models and providing recourse, empowering people to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "loan application accepted"). In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual; CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.

### 2025-06-30 paper/sections/method.qmd
Fix some awkward references after reformatting for AAAI

@grathwohl2020your observed that any classifier can be re-interpreted as a joint energy-based model that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. @grathwohl2020your show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.

@@

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE. One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.

>>>>

@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.

@@

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we work with a different distinction between CE and AE than @freiesleben2022intriguing who considers misclassification as the distinguishing feature of adversarial examples. One of the key observations of this work is that we can leverage CEs during training and get AEs essentially for free to reap the aforementioned benefits of adversarial training.
